---
title: "Analyzing NBA Player Performance Over Time"
author: "Daniela Nieto, Leopoldo Nikonowicz"
date: "August 3, 2023"
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: yes
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Introduction:

The data file contains per-game statistics of NBA players for every season from 1949 to
2023. The dataset includes essential variables such as:

 - Player: The name of the player
 - Team Name: The name of the NBA team the player is a part of.
 - Ranking: The player’s ranking during the respective season.
 - Points: The total points scored by the player during the season.
 - Field Goals (FG): The number of successful field goals made by the player.
 - Field Goal Attempts (FGA): The total number of field goal attempts by the player.
 - 2-Point Field Goals (2P): The number of successful two-point field goals made.
 - 2-Point Field Goal Attempts (2PA): The total number of two-point field goal attempts.
 - Free Throws (FT): The number of successful free throws made.
 - Assists (AST): The total assists made by the player.
 - Fouls (FL): The total number of fouls committed by the player.
 - Rebounds (REB): The total number of rebounds grabbed by the player.

The NBA per-game statistics dataset is a valuable resource for understanding players
performance and trends over the past decades. The data needs to be collected from
sports databases. The source for the NBA per-game statistics dataset is the following:

 - https://www.basketball-reference.com/leagues/NBA_2023_per_game.html

We will gather per-game statistics from the cited source covering seasons from 1946 to
2023. The data needs to be compiled and a new column for the corresponding season
year will be added to facilitate analysis. The dataset will contain a total of 31 variables.
There are about 200-600 players per season. Once the data is compiled, we will be able
to meet the 2,000-observation threshold.

Our interest in this dataset is driven by our passion for sports, particularly basketball,
and our curiosity about how players performances have evolved over time. By analyzing
the data, we aim to gain insights into the patterns and correlations that determine
players performance. We believe that understanding the factors influencing rankings
can provide valuable insights for coaches, analysts, and team managers in optimizing
strategies, player selection, and overall team development.

## Methods:

```{r, message = FALSE, warning = FALSE}
library(faraway)
library(lmtest)
library(leaps)
library(ggplot2)
library(readr)
```

### Functions:

```{r}
plot_model = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  par(mfrow = c(1, 2))
  plot(fitted(model), resid(model), 
       col = pointcol, pch = 20, cex = 1.5,
       xlab = "Fitted", ylab = "Residuals")
  abline(h = 0, col = linecol, lwd = 2)
  
  qqnorm(resid(model), col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)  
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```


### Data Loading and Cleaning:

*Load data and Modified names*

Utilizing the set of variables within the dataset would often confuse R; therefore, the names were adjusted for enhanced usability.
*Modified names*
```{r, message = FALSE}
nba = read_csv("NBA_data.csv")
names(nba)[names(nba) == "3P"] = "ThreeP"
names(nba)[names(nba) == "3PA"] = "ThreePA"
names(nba)[names(nba) == "3P%"] = "ThreePP"

names(nba)[names(nba) == "2P"] = "TwoP"
names(nba)[names(nba) == "2PA"] = "TwoPA"
names(nba)[names(nba) == "2P%"] = "TwoPP"

names(nba)[names(nba) == "FG%"] = "FGP"
names(nba)[names(nba) == "eFG%"] = "eFGP"
names(nba)[names(nba) == "FT%"] = "FTP"

```

The purpose of this was to assess the data for collinearity issues, with additional examples provided in the appendix.

```{r}
three_pointers = nba[, c("ThreeP", "ThreePA", "ThreePP")]
pairs(three_pointers)
```


Due to the presence of several missing values, the rows containing such data were removed from the set. The data showing collinearity issues where
also removed from the set.

```{r}

nba = nba[, -which(names(nba) == "Player"| names(nba) == "Pos" | names(nba) == "Tm" |
                      names(nba) == "FGP" | names(nba) == "FGA" | names(nba) == "ThreePP" | names(nba) == "ThreePA" | 
                      names(nba) == "TwoPP" | names(nba) == "TwoPA" | names(nba) == "FTP"|
                      names(nba) == "FTA" | names(nba) == "ORB" | names(nba) == "DRB")]

# Remove rows with missing values
cleaned_nba = na.omit(nba)
cleaned_nba = cleaned_nba[cleaned_nba$PTS != 0, ]
cleaned_nba = cleaned_nba[cleaned_nba$eFGP < 1 & cleaned_nba$eFGP > 0, ]

```

### Models:

*BACK BIC*
```{r, message = FALSE}
# Create large model
large_model = lm(PTS ~ ., data = cleaned_nba)

n = nrow(cleaned_nba)

# Perform backward stepwise selection
back_bic = step(large_model, k = log(n), trace = 0)
vif(back_bic)
back_bic = lm(PTS ~ G + FG + FT, data = cleaned_nba)
vif(back_bic)
```

*Interaction Model*
```{r}
int_model = lm(PTS ~ G * FG * FT, data = cleaned_nba)
```


## Results

```{r}
plot_model(back_bic)
```

```{r}
get_loocv_rmse(back_bic)
get_adj_r2(back_bic)
get_num_params(back_bic)
```

```{r}
plot_model(int_model)
```

```{r, eval = FALSE}
get_loocv_rmse(int_model)
get_adj_r2(int_model)
get_num_params(int_model)
```

```{r}
anova(back_bic, int_model)
```


## Discussion

Based on the anova test we prefer the interaction model. But 


## Appendix

```{r, message = FALSE}
two_pointers = nba[, c("TwoP", "TwoPA", "TwoPP")]
pairs(two_pointers)
```

```{r, message = FALSE}
free_throw = nba[, c("FT", "FTA", "FTP")]
pairs(free_throw)
```

```{r, message = FALSE}
field_goals = nba[, c("FG", "FGA", "FGP", "eFGP")]
pairs(field_goals)
```

```{r, message = FALSE}
rebounds = nba[, c("ORB", "DRB", "TRB")]
pairs(rebounds)
```

```{r, message = FALSE}
percentage = nba[, c("FGP", "ThreePP", "TwoPP", "eFGP", "FTP", "TRB")]
pairs(percentage)
```

```{r, message = FALSE}
games = nba[, c("G", "GS", "MP")]
pairs(games)
```
