---
title: "Untitled"
output: html_document
date: "2023-07-28"
---

```{r, message = FALSE, warning = FALSE}
library(faraway)
library(lmtest)
library(leaps)
library(ggplot2)
library(readr)
```

*Load data*
```{r, message = FALSE}
nba = read_csv("NBA_data.csv")
str(nba)
```
*Modified names*
```{r, message = FALSE}
names(nba)[names(nba) == "3P"] = "ThreeP"
names(nba)[names(nba) == "3PA"] = "ThreePA"
names(nba)[names(nba) == "3P%"] = "ThreePP"

names(nba)[names(nba) == "2P"] = "TwoP"
names(nba)[names(nba) == "2PA"] = "TwoPA"
names(nba)[names(nba) == "2P%"] = "TwoPP"

names(nba)[names(nba) == "FG%"] = "FGP"
names(nba)[names(nba) == "eFG%"] = "eFGP"
names(nba)[names(nba) == "FT%"] = "FTP"
```

```{r, message = FALSE}
three_pointers = nba[, c("ThreeP", "ThreePA", "ThreePP")]
pairs(three_pointers)
```

```{r, message = FALSE}
two_pointers = nba[, c("TwoP", "TwoPA", "TwoPP")]
pairs(two_pointers)
```

```{r, message = FALSE}
free_throw = nba[, c("FT", "FTA", "FTP")]
pairs(free_throw)
```

```{r, message = FALSE}
field_goals = nba[, c("FG", "FGA", "FGP", "eFGP")]
pairs(field_goals)
```

```{r, message = FALSE}
rebounds = nba[, c("ORB", "DRB", "TRB")]
pairs(rebounds)
```

```{r, message = FALSE}
percentage = nba[, c("FGP", "ThreePP", "TwoPP", "eFGP", "FTP", "TRB")]
pairs(percentage)
```

```{r, message = FALSE}
games = nba[, c("G", "GS", "MP")]
pairs(games)
```


```{r, echo = FALSE}
plot_fitted_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  plot(fitted(model), resid(model), 
       col = pointcol, pch = 20, cex = 1.5,
       xlab = "Fitted", ylab = "Residuals")
  abline(h = 0, col = linecol, lwd = 2)
}

plot_qq = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  qqnorm(resid(model), col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)
}
```

```{r, message = FALSE, warning = FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```


Rk -- Rank
Pos -- Position
Age -- Player's age on February 1 of the season
Tm -- Team
G -- Games
GS -- Games Started
MP -- Minutes Played Per Game
FG -- Field Goals Per Game
FGA -- Field Goal Attempts Per Game
FG% -- Field Goal Percentage
3P -- 3-Point Field Goals Per Game
3PA -- 3-Point Field Goal Attempts Per Game
3P% -- 3-Point Field Goal Percentage
2P -- 2-Point Field Goals Per Game
2PA -- 2-Point Field Goal Attempts Per Game
2P% -- 2-Point Field Goal Percentage
eFG% -- Effective Field Goal Percentage
This statistic adjusts for the fact that a 3-point field goal is worth one more point than a 2-point field goal.
FT -- Free Throws Per Game
FTA -- Free Throw Attempts Per Game
FT% -- Free Throw Percentage
ORB -- Offensive Rebounds Per Game
DRB -- Defensive Rebounds Per Game
TRB -- Total Rebounds Per Game
AST -- Assists Per Game
STL -- Steals Per Game
BLK -- Blocks Per Game
TOV -- Turnovers Per Game
PF -- Personal Fouls Per Game
PTS -- Points Per Game


```{r, message = FALSE}
# Remove columns
nba = read_csv("NBA_data.csv")
names(nba)[names(nba) == "3P"] = "ThreeP"
names(nba)[names(nba) == "3PA"] = "ThreePA"
names(nba)[names(nba) == "3P%"] = "ThreePP"

names(nba)[names(nba) == "2P"] = "TwoP"
names(nba)[names(nba) == "2PA"] = "TwoPA"
names(nba)[names(nba) == "2P%"] = "TwoPP"

names(nba)[names(nba) == "FG%"] = "FGP"
names(nba)[names(nba) == "eFG%"] = "eFGP"
names(nba)[names(nba) == "FT%"] = "FTP"
nba = nba[, -which(names(nba) == "Player"| names(nba) == "Pos" | names(nba) == "Tm" |
                      names(nba) == "FGP" | names(nba) == "FGA" | names(nba) == "ThreePP" | names(nba) == "ThreePA" | 
                      names(nba) == "TwoPP" | names(nba) == "TwoPA" | names(nba) == "FTP"|
                      names(nba) == "FTA" | names(nba) == "ORB" | names(nba) == "DRB")]

# Remove rows with missing values
cleaned_nba = na.omit(nba)
cleaned_nba = cleaned_nba[cleaned_nba$PTS != 0, ]
cleaned_nba = cleaned_nba[cleaned_nba$eFGP < 1 & cleaned_nba$eFGP > 0, ]
cleaned_nba$eFGP = cleaned_nba$eFGP*100
```

*BACK BIC*
```{r, message = FALSE}
# Create large model
large_model = lm(PTS ~ ., data = cleaned_nba)

n = nrow(cleaned_nba)

# Perform backward stepwise selection
back_bic = step(large_model, k = log(n), trace = 0)
vif(back_bic)
back_bic = lm(PTS ~ G + FG + FT, data = cleaned_nba)
vif(back_bic)
```

```{r, echo = FALSE}
plot_fitted_resid(back_bic)
plot_qq(back_bic)
```

```{r}
get_loocv_rmse(back_bic)
get_adj_r2(back_bic)
get_num_params(back_bic)
```

```{r}
int_model = lm(PTS ~ G * FG * FT, data = cleaned_nba)
```


```{r, echo = FALSE}
plot_fitted_resid(int_model)
plot_qq(int_model)
anova(back_bic, int_model)
```

```{r, eval = FALSE}
get_loocv_rmse(int_model)
get_adj_r2(int_model)
get_num_params(int_model)
```
